---
title: "Presidential Approval Rating"
output:
  html_document:
    df_print: paged
---

**Tea Break R**

**A quick tutorial on Presidential Approval Rating**

So today (Nov 6th, 2018) is Midterm Election Day. I can't think of a more fitting event to do a quick R tutorial to show you how to visualize Presidential Approval Rating with just a few lines of code. 

In this tutorial, I will show you how to compare the Presidential Approval Rating for President Trump and Obama during the first two years of their Presidency. I obtained data from the American Presidency Project. 

All you have to do is to download the two datasets for President Trump and President Obama. The approval rating shows the percentage of respondents to the opinion poll conducted by Gallup who approve of the sitting President. 

#### Data sources
https://docs.google.com/spreadsheets/d/1iEl565M1mICTubTtoxXMdxzaHzAcPTnb3kpRndsrfyY/edit?ts=5bd7f609#gid=671375968

**References**

+  https://news.gallup.com/interactives/185273/presidential-job-approval-center.aspx
+  https://www.presidency.ucsb.edu/statistics/data/presidential-job-approval
+  https://en.wikipedia.org/wiki/United_States_presidential_approval_rating

#### Load the necessary libraries

```{r}
library(tidyverse) # For data wrangling and data manipulation
library(lubridate) # For date manipulation
```

##### Read in President Trump's approval rating for the first two years and take a look at the data structure
```{r}
trump = read.csv('trump_approval.csv')

# Data structure
glimpse(trump)
```

##### Format the dates (currently in 'factor' format) and arrange the dates in ascending order
```{r}
trump$Start.Date = mdy(trump$Start.Date) # change 'Start.Date' to 'Date' format
trump$End.Date = mdy(trump$End.Date) # change 'End.Date' to 'Date' format

trump = trump %>%
  arrange(-desc(Start.Date)) # arrange dates in ascending order

glimpse(trump) # take a look at the new dataset
```

##### Take a look at the summary of our dataframe
```{r}
summary(trump)
```

##### Read in President Obama's approval rating for the first two years and take a look at the data structure. Since President Obama was in office for 8 years, we will filter the dataset to look at the first two years he was in office to make it more comparable to the dataset for President Trump

```{r}
obama_full = read.csv('obama_approval.csv') # read the dataset in 

obama_full$Start.Date = mdy(obama_full$Start.Date) # format date-related columns
obama_full$End.Date = mdy(obama_full$End.Date) # format date-related columns

glimpse(obama_full) # take a look at the data structure
```

##### Format the date-related columns of 'obama_full' and filter it to get data for the first two years only (2009 - 2010). We can see that the 'trump' dataframe has data recorded every 7 days starting from '2017-01-23.' Therefore, let's filter 'obama_full' dataframe based on the same structure (first two years and data recorded every 7 days) 

```{r}
obama_full = obama_full %>%
        arrange(-desc(Start.Date))%>% # sort data based on 'Start.Date' ascendingly
        filter(Start.Date >= ymd('2009-01-23') & Start.Date <= ymd('2010-10-25'))

summary(obama_full)
```

##### Take a look at the data structure again
```{r}
glimpse(obama_full)
```

##### Select the data recorded every 7 days for put them in a new dataframe called 'obama.' This will be comparable to the 'trump' dataframe
```{r}
obama = obama_full[seq(1, nrow(obama_full), 7),] # get data for every 7 days

rownames(obama) = seq(length = nrow(obama)) # reset index
```

##### Now, we are ready to make our first plot. Compare the average approval rating during the first two years in office of Obama and Trump. The process is as follows:

+  Extract the mean approval rating for each president
+  Combine the two dataframes
+  Make a bar plot

**Extract the average approval rating for Obama**
```{r}
obama_mean_approval = 
  obama %>% 
  select(Start.Date, Approving) %>%
  summarize(mean_approval = mean(Approving))

obama_mean_approval
```

**Extract the average approval rating for Trump**
```{r}
trump_mean_approval = 
  trump %>% 
  select(Start.Date, Approving) %>%
  summarize(mean_approval = mean(Approving))

trump_mean_approval
```

**Combine 'obama_mean_approval' and 'trump_mean_approval'**
```{r}
mean_approval = rbind(obama_mean_approval, trump_mean_approval) # bind the two dataframes by row

mean_approval$President = c('Obama', 'Trump') # rename the two columns

glimpse(mean_approval) # take a look at the data structure of the new dataframe
```

**Plot a bar plot to compare the average approval rating for two Presidents**

```{r}
ggplot(mean_approval, 
       aes(x = President, y = mean_approval)) +
  geom_bar(stat = 'identity', width = 0.4, fill = 'chocolate') + # set width & color
  ggtitle('Average Presidential Approval Rating (%) \n after the First Two Years in Office') + # add title
  coord_flip() + # change vertical bar plot to horizontal bar plot
  theme_minimal()+ # choose the minimal theme
  theme(axis.title.x = element_blank(), # remove x axis title
  axis.title.y = element_blank(),  # remove y axis title
  plot.title = element_text(color = 'chocolate', size = 15, face = 'bold', hjust = 0.5), # set color and size for title
  axis.text.y = element_text(color = 'chocolate', size = 15, face = 'bold'), # set color and size for texts on y axis
  axis.text.x = element_text(color = 'chocolate', size = 15, face = 'bold'), # set color and size for texts on x axis
  panel.grid.major = element_blank(), # remove major grid lines
  panel.grid.minor = element_blank()) + # remove minor grid lines
  geom_label(aes(label = round(mean_approval,1)), size = 5) # add labels (average approval rating score)
```

**Obama's average approval rating is 52.8% while Trump's average approval rating is 39.4%. It will be interesting to look at the fluctuation in approval rating of the two Presidents over the first two years. Let's proceed with our second plot.**

##### For our second plot, we are interested in plotting two line plots indicating Trump's and Obama's approval rating over the first two years. We will proceed with the following steps:

+  Rename the columns in two dataframes: 'obama' and 'trump'
+  Combine the two dataframes
+  Plot the two line plots

###### Step 1: Rename the columns
```{r}
colnames(obama) = c('o_start_date', 'o_end_date', 'o_approving',
                    'o_disapproving', 'o_no_data')

colnames(trump) = c('t_start_date', 't_end_date', 't_approving',
                    't_disapproving', 't_no_data')
```

**Notice we have 89 observations for 'obama' and 90 observations for 'trump.' Let's remove the last observation in the 'trump' dataframe**

```{r}
trump = trump[c(1:89),]
```

###### Step 2: Combine the two dataframes

```{r}
obama_trump = cbind(obama, trump) # combine two dataframes


glimpse(obama_trump) # take a look at the data structure
```


###### Step 3: Plot two line plots 

```{r}
ggplot(obama_trump) + # data frame
  geom_line(aes(o_start_date, y = o_approving), color = 'blue') + # Obama line
  geom_line(aes(o_start_date, y = t_approving), color = 'red') + # Trump line
  ggtitle('Weekly Average Presidential Approval Rating (%) during the First Two Years \n    January 2009 - October 2010 and January 2017 - October 2018') + # add title
  theme_minimal() + # set thme
    # set color, font, and size for title, texts for x axis, and texts for y axis
  theme(plot.title = element_text(size = 13, face = 'bold', hjust = 0.5),
        axis.text.y = element_text(size = 12, face = 'bold'),
        axis.text.x = element_text(size = 12, face = 'bold'),
        axis.title.y = element_blank(), # remove y-axis title
        axis.title.x = element_blank(), # remove x-axis title
        panel.grid.minor = element_blank()) + # remove minor grid
        scale_y_continuous(breaks = seq(0, 80, 5)) + # set y-scale # interval of 5 for easy representation
  scale_x_date(date_breaks = "2 month", # set date break every 2 months and extract the month only 
               date_labels = "%b") +
  annotate("text", x = as.Date('2010-10-08'), # annotation
           label = 'Obama', y = 50, size = 6, fontface = 'bold') +
  annotate("text", x = as.Date('2010-10-08'), # annotation
           label = 'Trump', y = 37, size = 6, fontface = 'bold')

```

There we have it. We can see that President Obama had a very high approval rating (more than 65%) at the start of his term and President Trump started off having only 45% approval rating. Obama's approval rating declined sharply within the first 6 months of his presidency and steadily declined thereafter. For President Trump, his approval rating has never been higher than 45%. Right before the Midterm Election, the two Presidents have very comparable approval rating: a little below 45%. 
