---
title: "Malaria Deaths"
output: html_notebook
---

#### This data exploratory project is part of TutorialTuesday

(https://www.linkedin.com/pulse/november-data-challenge-kate-strachnyi-/)

#### and TidyTuesday 

(https://github.com/rfordatascience/tidytuesday)

**Load libraries**

```{r}
library(tidyverse)
library(lubridate)
library(skimr)
library(RColorBrewer)
library(scales)
```

**Import datasets**

```{r}
malaria_deaths = read.csv("malaria_deaths_by_age.csv")

# https://www.gapminder.org/data/geo/
geography = read.csv("geography.csv")
```

**Explore our data**

```{r}
malaria_deaths %>% glimpse()
```

```{r}
geography %>% glimpse()
```

**Let's clean up the data and rename some of the columns**

```{r}
malaria_deaths$X = NULL
colnames(malaria_deaths) = c("Country", "Code", "Year", "Age_Group", "Deaths")
```

**Review our datasets before performing joining**
```{r}
summary(malaria_deaths)
```

**Return the number of countries in the "maria_deaths"**
```{r}
malaria_deaths %>% 
  select(Country) %>%
  unique() %>% 
  count()
```

**Make the "Code" column lowercase. We do this as eventually we want to merge this one with the "geography" dataset**
```{r}
malaria_deaths$Code = tolower(malaria_deaths$Code)
```

**Return the number of countries in "geography"**
```{r}
geography %>% select(name) %>% count() 
```

**More data manipulation for "geography"**
```{r}
colnames(geography)[1:2] <- c("Code", "Country")

glimpse(geography)
```

**Inner join "malaria_deaths" onto geography**
```{r}
malaria_deaths = malaria_deaths %>%
  inner_join(geography, by = "Code")

colnames(malaria_deaths)
```

**Return some descriptive stats for the joined dataframe**
```{r}
summary(malaria_deaths)
```

**More data manipulation**
```{r}
colnames(malaria_deaths)[1] = "Country"

colnames(malaria_deaths)[15] = "Income_Group"

colnames(malaria_deaths)[7] = "Four_Reg"

colnames(malaria_deaths)[9] = "Six_Reg"
```

**Subset the joined dataset to get only the variables we want**
```{r}
malaria_deaths = 
  malaria_deaths[c("Country", "Year", "Age_Group", "Deaths", "Four_Reg", "Six_Reg",
                   "Latitude", "Longitude", "Income_Group", "Latitude", "Longitude")]
```

```{r}
summary(malaria_deaths)
```


**Plot a bar plot to show total number of deaths by age group for each region**
```{r}
ggplot(malaria_deaths) +
  geom_bar(stat = "identity", width = 1.5, aes(x = Four_Reg, y = Deaths, fill = Age_Group), position = "dodge") +
  scale_y_continuous(limits = c(0, 250000)) +
  ggtitle("Total Number of Malaria Deaths for Different Age Groups by Region (1990 - 2016)") +
  theme_minimal() +
  xlab("World Regions") +
  ylab("Number of Deaths") +
  theme(axis.text.y = element_text(size = 12, face = 'bold'),
        axis.text.x = element_text(size = 12, face = 'bold'),
        axis.title.y = element_text(size = 12, face = 'bold'),
        axis.title.x = element_text(size = 12, face = 'bold'),
        plot.title = element_text(size = 12, face = "bold")) +
  scale_fill_brewer(palette = "Set1")
```

**It seems Africa exhibits an extremely high number of deaths. To see the picture more clearly, let's look at Africa and the rest of the world separately**

```{r}
africa = 
  malaria_deaths %>%
  filter(Four_Reg == "africa")
```

```{r}
ggplot(africa) +
  geom_bar(stat = "identity", aes(x = Four_Reg, y = Deaths, fill = Age_Group), position = "dodge") +
  scale_y_continuous(limits = c(0, 250000)) +
  ggtitle("Total Number of Malaria Deaths for Different Age Groups in Africa (1990 - 2016)") +
  theme_minimal() +
  xlab("Africa") +
  ylab("Number of Deaths") +
  theme(axis.text.y = element_text(size = 12, face = 'bold'),
        axis.text.x = element_text(size = 12, face = 'bold'),
        axis.title.y = element_text(size = 12, face = 'bold'),
        axis.title.x = element_text(size = 12, face = 'bold'),
        plot.title = element_text(size = 12, face = "bold"),
        panel.grid.major = element_blank()) +
  scale_fill_brewer(palette = "Set1")
```

We can see that the group affected the most in Africa is the "Under 5" group. Now, let's look at the rest of the world. 

```{r}
world_no_africa = 
  malaria_deaths %>%
  filter(Four_Reg != "africa")
```

```{r}
ggplot(world_no_africa) +
  geom_bar(stat = "identity", width = 0.8, aes(x = Four_Reg, y = Deaths, fill = Age_Group), position = "dodge") +
  scale_y_continuous(limits = c(0, 50000)) +
  ggtitle("Total Number of Malaria Deaths for Different Age Groups by Region (1990 - 2016)") +
  theme_minimal() +
  xlab("World Regions") +
  ylab("Number of Deaths") +
  theme(axis.text.y = element_text(size = 12, face = 'bold'),
        axis.text.x = element_text(size = 12, face = 'bold'),
        axis.title.y = element_text(size = 12, face = 'bold'),
        axis.title.x = element_text(size = 12, face = 'bold'),
        plot.title = element_text(size = 12, face = "bold")) +
  scale_fill_brewer(palette = "Set1")
```

We can see that the death toll in Americas and Europe is close to zero. The group affected the most in Asia is also the "Under 5" group. 

**Now, let's look at the trend of the country with the most number of malaria deaths**

```{r}
malaria_deaths %>%
  arrange(desc(Deaths))
```

**The country with the highest number of malaria deaths over the year is Nigeria. Let's zoom into Nigeria and see the trend over the year**
```{r}
nigeria = 
  malaria_deaths %>%
  filter(Country == "Nigeria")

head(nigeria)
```

**Explore our new filtered dataset**
```{r}
glimpse(nigeria)
```

```{r}
summary(nigeria)
```

**Plot the number of malaria deaths in Nigeria over the period 1990 - 2016 for different age groups**
```{r}
ggplot(nigeria, aes(x = Year, y = Deaths)) +
  geom_line(aes(color = Age_Group)) +
  ggtitle("Total Number of Malaria Deaths for Different Age Groups in Nigeria (1990 - 2016)") +
  theme_minimal() +
  xlab("Year") +
  ylab("Number of Deaths") +
  theme(axis.text.y = element_text(size = 12, face = 'bold'),
        axis.text.x = element_text(size = 12, face = 'bold'),
        axis.title.y = element_text(size = 12, face = 'bold'),
        axis.title.x = element_text(size = 12, face = 'bold'),
        plot.title = element_text(size = 12, face = "bold", hjust = 0.2),
        panel.grid.major = element_blank()) +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(breaks = pretty_breaks(n = 10)) +
  scale_y_continuous(breaks = pretty_breaks(n = 5))

```

